Mathematical appendix to "Spatio-temporal estimates of HIV risk group proportions for adolescent

### 2022-07-04 src/docs_paper/appendix.Rmd
Small appendix fixes SRF #128

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table B.5) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$ and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.
We used a multinomial logistic regression model to infer the proportion of AGYW in the risk groups $k \in \{1, 2, 3^{+}\}$.
Typically, this model would be specified using a multinomial likelihood
\begin{equation}
    \y_{ita} = (y_{ita1}, \ldots, y_{ita3^{+}})^\top \sim \text{Multinomial}(m_{ita}; \, p_{ita1}, \ldots, p_{ita3^{+}}),
\end{equation}
where the number of women in risk group $k$ is $y_{itak}$, the sample size is $m_{ita} = \sum_{k = 1}^{3^{+}} y_{itak}$, and $p_{itak} > 0$ is the probability of membership of the $k$th risk group with $\sum_{k = 1}^{3^{+}} p_{itak} = 1$.
Following the baseline category approach with $k = 1$ say, linear predictors would then be specified for the additive log ratios $\log(p_{itak} / p_{ita1})$.
Instead of taking this approach, to facilitate inference in the `R-INLA`, we used an equivalent Poisson log-linear model via the multinomial-Poisson transformation [@baker1994multinomial].
This transformation, and further details about this model, are presented in Section 1.2.

Next, we fit a logistic regression model to estimate the proportion of those in the $k = 3^{+}$ risk group that were in the $k = 3$ and $k = 4$ risk groups respectively.
This model was of the form
\begin{align}
    y_{ia4} &\sim \text{Binomial} \left( y_{ia3} + y_{ia4}, q_{ia} \right), \\
    q_{ia} &= p_{ia4} / (p_{ia3} + p_{ia4}) = p_{ia4} / p_{ia{3^+}}, \\
    q_{ia} &= \text{logit}^{-1} \left( \eta_{ia} \right), 
\end{align}
for suitable choice of the linear predictor $\eta_{ia}$, described in Section 1.3.
Taking this modular approach allowed us to include all surveys in the multinomial regression model, but only those surveys with a specific transactional sex question to estimate the proportions $\{p_{ia4}\}$.
As all such surveys occurred in 2013-2018, in the logistic regresion model $q_{ia}$ was assumed to be constant over time.

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{lllllll}
\midrule
Model ID & Category ($\beta_k$) & Age ($\alpha_{ak}$) & Country ($\zeta_{ck}$) & Spatial ($\phi_{ik}$) & Temporal ($\gamma_{tk}$) & Spatio-temporal ($\delta{itk}$) \\
\midrule
M1 & IID & IID & IID & IID & IID & \xmark \\
M1x & IID & IID & IID & IID & IID & IID $\otimes$ IID \\
M2 & IID & IID & IID & Besag & IID & \xmark \\
M2x & IID & IID & IID & Besag & IID  & Besag $\otimes$ IID \\
M3 & IID & IID & IID & IID & AR1 & \xmark \\
M3x & IID & IID & IID & IID & AR1 & IID $\otimes$ AR1 \\
M4 & IID & IID & IID & Besag & AR1 & \xmark \\
M4x & IID & IID & IID & Besag & AR1 & Besag $\otimes$ AR1 \\
\midrule
\end{tabularx}
\caption{The multinomial regression models that we considered. Observation random effects $\theta_{ita}$ are omitted from this table.}
\label{tab:multinomial-models}
\small
\end{table}

We considered eight models (Table \ref{tab:multinomial-models}) for $\eta_{ita}$ of the form
$$
\eta_{ita} = \theta_{ita} + \beta_k + \alpha_{ak} + \zeta_{c[i]k} + \phi_{ik} + \gamma_{tk} + \delta_{itk}.
$$
As described above, we included observation random effects in all models we considered, and set $\theta_{ita} \sim \mathcal{N}(0, \tau_\theta^{-1})$, with fixed precision $\tau_\theta = 0.000001$ (or a standard deviation of $\sigma_\theta = 1000$).
To capture country-specific proportion estimates for each category, we included category random effects $\beta_k \sim \mathcal{N}(0, \tau_\beta^{-1})$ and country-category random effects $\zeta_{ck} \sim \mathcal{N}(0, \tau_\zeta^{-1})$.
Heterogeneity in risk group proportions by age was allowed by including age-country-category random effects $\alpha_{ack} \sim \mathcal{N}(0, \tau_\alpha^{-1})$.
We considered two specifications, independent and identically distributed (IID) and Besag, for the space-category $\phi_{ik}$ random effects (Section \ref{sec:spatial-random-effects}) and two specifications, IID and first order autoregressive (AR1), for the year-category $\gamma_{tk}$ random effects (Section \ref{sec:temporal-random-effects}).
We also considered separable space-year-category interactions $\delta_{itk}$ with corresponding Kronecker product structure (Section \ref{sec:spatio-temporal-random-effects}).

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{llllll}
\midrule
Model ID & Intercept ($\beta_0$) & Age ($\alpha_a$) & Country ($\zeta_{c}$) & Spatial ($\phi_i$) & Covariates \\ 
\midrule
L1 & Constant & IID & IID & IID & \xmark \\
L2 & Constant & IID & IID & Besag & \xmark \\
L3 & Constant & IID & IID & IID & \texttt{cfswever} \\
L4 & Constant & IID & IID & Besag & \texttt{cfswever} \\
L5 & Constant & IID & IID & IID & \texttt{cfswrecent} \\
L6 & Constant & IID & IID & Besag & \texttt{cfswrecent} \\
\midrule
\end{tabularx}
\caption{The logistic regression models that were considered. \texttt{cfswever} denotes the proportion of men who have ever paid for sex and \texttt{cfswrecent} denotes the proportion of men who have paid for sex in the past 12 months.}
\label{tab:logistic-models}
\small
\end{table}

@@

As with the multinomial regression model, we used survey weighted counts $\{y_{itak}^\star\}$ and sample sizes $\{m_{itak}^\star\}$.
We used a used a generalised binomial pseudo-likelihood $y^\star \sim \text{xBinomial}(y^\star, m^\star)$, as implemented by \texttt{family = "xBinomial"} in \texttt{R-INLA}, given by
\begin{equation}
    p(y^\star \, | \, m^\star) =  \binom{\lfloor m^\star \rfloor}{\lfloor y^\star \rfloor} q^{y^\star} (1 - q)^{m^\star - y^\star}.
\end{equation}
to extend the binomial distribution to non-integer weighted counts and sample sizes.

@@

\begin{table}[h]
\centering
\begin{tabularx}{0.82\textwidth}{lllllll}
\toprule
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\ 
\midrule
DIC & 4662 (110) & 4605 (111) & 4662 (110) & 4605 (111) & 4662 (110) & 4605 (111) \\ 
WAIC & 4692 (115) & 4624 (115) & 4692 (115) & 4624 (115) & 4692 (115) & 4624 (115) \\ 
CPO & 950 (15) & 969 (15) & 951 (15) & 970 (15) & 950 (15) & 970 (15) \\ 
\bottomrule
\end{tabularx}
\caption{Logistic regression model performance.}
\label{tab:logistic-model-performance}
\end{table}

\begin{figure}[h]
    \centering
    \makebox[\textwidth][c]{\includegraphics{depends/fsw-logit-information-criteria.jpeg}}
    \caption{Logistic regression model performance.}
    \label{fig:logistic-model-performance}
\end{figure}

>>>>

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table B.5) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$ and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$.
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.
We used a multinomial logistic regression model to infer the proportion of AGYW in the risk groups $k \in \{1, 2, 3^{+}\}$.
This model can be specified using a multinomial likelihood
\begin{equation}
    \y_{ita} = (y_{ita1}, \ldots, y_{ita3^{+}})^\top \sim \text{Multinomial}(m_{ita}; \, p_{ita1}, \ldots, p_{ita3^{+}}),
\end{equation}
where the number of women in risk group $k$ is $y_{itak}$, the sample size is $m_{ita} = \sum_{k = 1}^{3^{+}} y_{itak}$, and $p_{itak} > 0$ is the probability of membership of the $k$th risk group with $\sum_{k = 1}^{3^{+}} p_{itak} = 1$.
Following the baseline category approach with $k = 1$, linear predictors may then be specified for the additive log ratios $\log(p_{itak} / p_{ita1})$.
To facilitate inference in the `R-INLA`, we used an equivalent Poisson log-linear model via the multinomial-Poisson transformation [@baker1994multinomial].
This transformation, and further details about this model, are presented in Section 1.2.

Next, we fit a logistic regression model to estimate the proportion of those in the $k = 3^{+}$ risk group that were in the $k = 3$ and $k = 4$ risk groups respectively.
This model was of the form
\begin{align}
    y_{ia4} &\sim \text{Binomial} \left( y_{ia3} + y_{ia4}, q_{ia} \right), \label{eq:logistic-regression} \\
    q_{ia} &= \text{logit}^{-1} \left( \eta_{ia} \right), 
\end{align}
where $q_{ia} = p_{ia4} / (p_{ia3} + p_{ia4}) = p_{ia4} / p_{ia{3^+}}$ and the linear predictor $\eta_{ia}$ is chosen suitably, described in Section 1.3.
Taking this two-step approach allowed us to include all surveys in the multinomial regression model, but only those surveys with a specific transactional sex question in Equation \ref{eq:logistic-regression}.
As all such surveys occurred in 2013-2018, in the logistic regression model $q_{ia}$ was assumed to be constant over time.

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{lllllll}
\midrule
Model ID & Category ($\beta_k$) & Age ($\alpha_{ak}$) & Country ($\zeta_{ck}$) & Spatial ($\phi_{ik}$) & Temporal ($\gamma_{tk}$) & Spatio-temporal ($\delta{itk}$) \\
\midrule
M1 & IID & IID & IID & IID & IID & Not included \\
M1x & IID & IID & IID & IID & IID & IID $\otimes$ IID \\
M2 & IID & IID & IID & Besag & IID & Not included \\
M2x & IID & IID & IID & Besag & IID  & Besag $\otimes$ IID \\
M3 & IID & IID & IID & IID & AR1 & Not included \\
M3x & IID & IID & IID & IID & AR1 & IID $\otimes$ AR1 \\
M4 & IID & IID & IID & Besag & AR1 & Not included \\
M4x & IID & IID & IID & Besag & AR1 & Besag $\otimes$ AR1 \\
\midrule
\end{tabularx}
\caption{The multinomial regression models that we considered. Observation random effects $\theta_{ita}$ are omitted from this table.}
\label{tab:multinomial-models}
\small
\end{table}

We considered eight models (Table \ref{tab:multinomial-models}) for $\eta_{ita}$ of the form
$$
\eta_{ita} = \theta_{ita} + \beta_k + \alpha_{ak} + \zeta_{c[i]k} + \phi_{ik} + \gamma_{tk} + \delta_{itk}.
$$
As described above, we included observation random effects in all models we considered, and set $\theta_{ita} \sim \mathcal{N}(0, 1000^2)$.
To capture country-specific proportion estimates for each category, we included category random effects $\beta_k \sim \mathcal{N}(0, \tau_\beta^{-1})$ and country-category random effects $\zeta_{ck} \sim \mathcal{N}(0, \tau_\zeta^{-1})$.
Heterogeneity in risk group proportions by age was allowed by including age-country-category random effects $\alpha_{ack} \sim \mathcal{N}(0, \tau_\alpha^{-1})$.
We considered two specifications, independent and identically distributed (IID) and Besag, for the space-category $\phi_{ik}$ random effects (Section \ref{sec:spatial-random-effects}) and two specifications, IID and first order autoregressive (AR1), for the year-category $\gamma_{tk}$ random effects (Section \ref{sec:temporal-random-effects}).
We also considered separable space-year-category interactions $\delta_{itk}$ with corresponding Kronecker product structure (Section \ref{sec:spatio-temporal-random-effects}).

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{llllll}
\midrule
Model ID & Intercept ($\beta_0$) & Age ($\alpha_a$) & Country ($\zeta_{c}$) & Spatial ($\phi_i$) & Covariates \\ 
\midrule
L1 & Constant & IID & IID & IID & Not included \\
L2 & Constant & IID & IID & Besag & Not included \\
L3 & Constant & IID & IID & IID & \texttt{cfswever} \\
L4 & Constant & IID & IID & Besag & \texttt{cfswever} \\
L5 & Constant & IID & IID & IID & \texttt{cfswrecent} \\
L6 & Constant & IID & IID & Besag & \texttt{cfswrecent} \\
\midrule
\end{tabularx}
\caption{The logistic regression models that were considered. \texttt{cfswever} denotes the proportion of men who have ever paid for sex and \texttt{cfswrecent} denotes the proportion of men who have paid for sex in the past 12 months.}
\label{tab:logistic-models}
\small
\end{table}

@@

As with the multinomial regression model, we used survey weighted counts $\{y_{itak}^\star\}$ and sample sizes $\{m_{itak}^\star\}$.
We used a generalised binomial pseudo-likelihood $y^\star \sim \text{xBinomial}(y^\star, m^\star)$, as implemented by \texttt{family = "xBinomial"} in \texttt{R-INLA}, given by
\begin{equation}
    p(y^\star \, | \, m^\star) =  \binom{\lfloor m^\star \rfloor}{\lfloor y^\star \rfloor} q^{y^\star} (1 - q)^{m^\star - y^\star}.
\end{equation}
to extend the binomial distribution to non-integer weighted counts and sample sizes.

@@

\begin{table}[h]
\centering
\begin{tabularx}{0.82\textwidth}{lllllll}
\toprule
 & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 \\ 
\midrule
DIC & 4662 (110) & 4605 (111) & 4662 (110) & 4605 (111) & 4662 (110) & 4605 (111) \\ 
WAIC & 4692 (115) & 4624 (115) & 4692 (115) & 4624 (115) & 4692 (115) & 4624 (115) \\ 
CPO & 950 (15) & 969 (15) & 951 (15) & 970 (15) & 950 (15) & 970 (15) \\ 
\bottomrule
\end{tabularx}
\caption{Logistic regression model performance. For the CPO, higher values indicate better model performance for the CPO. For the DIC and WAIC, lower values indicate better model performance.}
\label{tab:logistic-model-performance}
\end{table}

\begin{figure}[h]
    \centering
    \makebox[\textwidth][c]{\includegraphics{depends/fsw-logit-information-criteria.jpeg}}
    \caption{Logistic regression model performance, with the best performing model(s) according to each criteria shown as a square.}
    \label{fig:logistic-model-performance}
\end{figure}
